{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-05 thesis/landslide.tex
handle most remaining todos in landslide

@@

						emit a preemption trace (\cref{sec:landslide-foundabug})

@@

						% may be informed by heuristix/yield blocking

@@

% TODO: add some backup linkeroos
The Bochs port of Landslide is likewise open-source
\revisionminor{under the \landslidelicense license}
and available at \url{https://github.com/bblum/landslide}.
% TODO: refresh
The HEAD commit at the time of writing is 8127151.
The Simics port uses Simics's proprietary API and is hence unlicensed and available upon request for educational use only.
Development on the Simics port is largely frozen,
as the Bochs port implements all the same features and more,
and is also roughly 3x faster.

@@

This section describes the features of Landslide the average student user should expect to interact with.
% TODO: make more specific section reference
Separate user guides also exist, described in \Cref{chap:education}.

@@

The preferred method of invoking Landslide is through Quicksand, the Iterative Deepening wrapper program which has all of \Cref{chap:quicksand} to itself.
% TODO: add quicksand symlink
This is done via the {\tt ./landslide} script in the top-level directory, which:
\begin{itemize}
	\item Checks if the user needs to run {\tt *-setup.sh} again, in case her source code was more recently updated than the existing annotated build (a common mistake),
	\item Passes its arguments through to {\tt id/landslide-id}, the Quicksand binary,
		and
	\item (If during the student user study,) compresses the resulting log files,
		creates a snapshot tarball of them and the current version of the user's code,
		and sends it to me for nefarious research purposes
		\revisionminor{(see \cref{sec:education-pebbles-instrumentation})}.
\end{itemize}

@@

\begin{itemize}
	\item {\tt -C}: enable ``control experiment'' mode, i.e., run only 1 instance of Landslide, with all (non-data-race) preemption points enabled in advance
		% TODO: put a section reference here
	\item {\tt -I}: enable Iterative Context Bounding (requires {\tt -C}, although future work may relax this restriction);
		this generally causes bugs to be found faster should they exist, but degrades completion time
		(\cref{sec:landslide-icb})
	\item {\tt -0}: enable Preempt-Everywhere mode (\cref{sec:quicksand-eval}, requires {\tt -C})
	\item {\tt -M}: enable Maximal State Space mode,
		which prioritizes the maximal state space to optimize for fast verification,
		abandoning all subset jobs even if they might find bugs faster
		(\cref{sec:tm-eval}, incompatible with {\tt -C}).
		According to \cref{sec:quicksand-soundness}'s soundness proofs,
		this is equivalent to {\tt -0}
		(and according to my experience, \revisionminor{much} faster as well).
	\item {\tt -H}: use Limited Happens-Before for data-race analysis (\cref{sec:background-hb})
		(default for Pebbles kernelspace mode)
	\item {\tt -V}: use vector-clock-based Pure Happens-Before for data-race analysis (\cref{sec:background-hb})
		(default for P2/PSU userspace and Pintos modes)
	\item {\tt -X}: support transactional memory (\Cref{chap:tm})
	\item {\tt -A}: support multiple abort codes during transaction failure (\cref{sec:tm-implementation});
		required for testing programs which behave differently under different abort circumstances,
		but impacts the state space size
	\item {\tt -S}: suppress retry aborts during transaction failure (\cref{sec:tm-implementation})
	\item {\tt -R}: enable retry-set state space reduction for transactional tests (\cref{sec:tm-implementation})
	\item {\tt -P}: support Pintos architecture (enabled automatically when {\tt pintos-setup.sh} is run)
	\item {\tt -4}: support Pebbles architecture (enabled automatically when either {\tt p2-setup.sh} or {\tt psu-setup.sh} is run)
		% TODO: put a section reference
	\item {\tt -e ETAFACTOR}: configure heuristic state space ETA deferring factor (described in detail in {\tt id/option.c})
	\item {\tt -E ETATHRESH}: configure heuristic threshold of state space progress for judging ETA stability (described in detail in {\tt id/option.c})
\end{itemize}

@@

\begin{itemize}
	\item {\tt within\_function FUNC} - adds {\tt FUNC} to an allowlist of functions required to appear in the current stack trace before identifying a preemption point (see \cref{sec:landslide-pps})
	\item {\tt without\_function FUNC} - as above, but a denylist instead of an allowlist
	\item {\tt within\_user\_function FUNC} - as two above but finds the function in the userspace test program rather than the kernel code.
	\item {\tt without\_user\_function FUNC} - difference to two above same as stated one above.
	\item {\tt data\_race ADDR TID LAST\_CALL CURRENT\_SYSCALL} - specifies a data-race preemption point.
		\begin{itemize}
			\llitem {\tt ADDR} shall be the code address (in hex) of the racing address,
			{\em before} the execution of which a preemption will be issued.
			\llitem {\tt TID} indicates a thread ID required to be running for this data race.
				To specify data race PPs across all threads at once, set {\tt FILTER\_DRS\_BY\_TID=0} (see next section).
			\llitem {\tt LAST\_CALL} indicates a code address required to be the site of the last {\tt call} instruction executed
				(similar to specifying a stack trace, but using a full stack trace here degrades performance too much),
				or 0 to not use this feature.
				From personal experience I found this option rather useless and recommend always supplying 0.
				% TODO: fix this section ref
				For further discussion see \cref{sec:quicksand-pps}.
			\llitem {\tt CURRENT\_SYSCALL} indicates the system call number if a user-space data race comes from within a kernel system call which accesses user memory (Pebbles only).
				Usually 0 (i.e., not in kernel code) but {\tt deschedule}'s system call number is common as well.
		\end{itemize}
	\item {\tt input\_pipe FILENAME} - FIFO file used for receiving messages from Quicksand (e.g. to suspend or resume execution).
		Requires {\tt id\_magic} option to be set (next section below).
		The odds that a human user will find spiritual enlightenment through using this option by hand are infinitesimal.
	\item {\tt output\_pipe FILENAME} - as above but for sending messages.
\end{itemize}

@@

\section{Kernel annotations}

@@

Hence,}
the educational experiments in this thesis focus on projects which students implement
on top of provided kernel basecode which Landslide already ``understands''.
Such understanding is conferred via the annotations described in this section.
For P2 and Pintos students I supply these annotations behind the scenes,
but a CMU 15-410 student who wishes to use Landslide on her kernel project shall need to brave forth hereupon.

@@

Traversal of the state space is implemented in three parts:
first, identifying preemption points when first encountered and selecting which thread to run for its first execution, in {\tt arbiter.c},
second, selecting which preemption point to backtrack to after completing an execution
and which thread to ``have switched to'' %
% TODO: post-revisions; uncomment; change " %" above to "%"
%\footnote{willan on-having switched to \cite{hhgg-reu}}
instead, in {\tt explore.c},
and third, rewinding the machine state to implement said backtracking,
in {\tt timetravel.c} (Bochs version) and {\tt timetravel-simics.c} (Simics version).

@@

The arbiter (named after the corresponding component of dBug \cite{dbug-ssv})
is responsible for checking which code locations during execution should be identified as preemption points
({\tt arbiter\_interested()}),
and thereupon for choosing whether to keep running the current thread or to preempt and switch to a new one
({\tt arbiter\_choose()}).
Its behaviour in the former case is configured by the options listed in \cref{sec:landslide-dynamicconfig},
and in the latter case by the options listed in \cref{sec:landslide-staticconfig}.
For example, {\tt EXPLORE\_BACKWARDS} is interpreted here;
if set, it will cause Landslide to always preempt and switch threads the first time it encounters each new preemption point.%
\footnote{Another secret option, {\tt CHOOSE\_RANDOMLY},
also exists here to randomize whether to ``explore backwards''
(choosing independently at each preemption point, resulting in an overall unpredictable exploration order).
% TODO: decide whether you're gonna put those estimate graphs in here
%% I
It's not exposed to {\tt config.landslide} but rather the user must edit it in {\tt arbiter.c} directly,
whereupon the probability may also be adjusted via {\tt numerator} and {\tt denominator}.}

@@

The {\tt data\_race} command specifies an instruction pointer value to identify as a data race preemption point.
The point can optionally be qualified by a thread ID, most recent system call number, et cetera,
as described in \cref{sec:landslide-dynamicconfig},
and is queried through {\tt suspected\_data\_race()}.

In Preempt-Everywhere mode, instead Landslide marks all shared memory accesses
as long as they are not either part of the mutex implementation or part of the running thread's stack frame.
The {\tt data\_race} command is ignored and {\tt suspected\_data\_race()}
instead checks whether the instruction pointer is associated with any such shared memory access.

@@

{\bf Choosing which new interleaving to test next.}
Among all interleavings chosen by DPOR not already marked as explored,
Landslide chooses the one with the longest execution prefix matching the current $S$,
to maintain the depth-first search invariant.
(In the case of a tie, differing only by which thread to run, it chooses arbitrarily.)
All other new interleavings are marked to explore later in their corresponding {\tt struct hax},
and automatically included in the result of any future iterations of DPOR until they are tested.
Because the one with the longest execution prefix was chosen to test next,
all others must share their execution prefixes with it,
preserving the $O(nk)$ memory bound described above.
%so storing markers for the others during the next execution is still $O(1)$ memory.

@@

In this interleaving, DPOR identifies 3 memory conflicts, two read/write and one write/write,
among the threads' 4 accesses to {\tt x}.
For each such pair, it ``marks'' an alternate interleaving,
which shall begin by preempting the thread of the first half of the conflict just before its execution thereof.
The ultimate goal is to execute an interleaving which reorders the conflict, which may expose new behaviour.
These marked interleavings form a work-queue which defines the exploration.
DPOR consumes from this set in depth-first order
(note the reversed order of $\bigstar$1 and $\bigstar$2)
to avoid storing in memory any representation of exponentially-sized subtrees outside of the current branch.
Note also that in $\bigstar$2,
the reordered \dporTB{{\tt tmp2~<-~x}} is not directly part of the memory conflict which marked it,
but it must be executed first to reach the conflicting \dporTB{{\tt x~<-~tmp2}}.

Now, let us run multiple iterations of DPOR
to advance through the first half of the full state space shown in \Cref{fig:tree}(b).
\Cref{fig:dpor-example-1} shows the outcome
(with the previously-marked $\bigstar$2, now re-labeled as $\bigstar$3, having its subtree abbreviated).

@@

% TODO: talk abt free re malloc false positives

@@

First of all, note that \Cref{alg:icb} is structured in a way that repeats interleavings
with fewer than $n$ preemptions that have already been checked in previous iterations of the outer loop.
This is because the number of preemptions in each branch is not known in advance;
rather, the state space must always be explored in an overall depth-first approach,
at best skipping too-preemptful interleavings as they are encountered.
As simple as it would be to state ``{\bf foreach} $S \in {\mathsf{sort}_P(U)}$'' in pseudocode,
implementing such an ordering would be much less straightforward. %
% TODO: post committee review
%\footnote{Translating conference papers' pseudocode into usable implementations is often quite difficult \cite{itg2}.}

>>>>

		Further interpret test-specific annotations
		as informed by
		configuration options (\cref{sec:landslide-directly}),
		test case annotations (\cref{sec:landslide-testcases}),
		and
		kernel annotations (\cref{sec:landslide-kernel-anno}).

@@

						emit a preemption trace
						(\cref{sec:landslide-foundabug}, \cref{sec:landslide-bugreport})

@@

@@

The Bochs port of Landslide is likewise open-source
\revisionminor{under the \landslidelicense license}
and available at \url{https://github.com/bblum/landslide}.
% TODO: refresh
The HEAD commit at the time of writing is 8127151.
In the event of a GitHub outage or the above link otherwise not working,
a backup of the repository
% checked out at the above commit
may be found at
% TODO: make there actually be smth at this commit
\url{https://www.cs.cmu.edu/~410/landslide.tar.bz2}.
%
The Simics port uses Simics's proprietary API and is hence unlicensed and available upon request for educational use only.
Development on the Simics port is largely frozen,
as the Bochs port implements all the same features and more,
and is also roughly 3x faster.

@@

This section describes the features of Landslide the average student user should expect to interact with.
Separate user guides also exist, described in \cref{sec:education-pebbles-recruiting}.

@@

The preferred method of invoking Landslide is through Quicksand, the Iterative Deepening wrapper program which has all of \Cref{chap:quicksand} to itself.
This is done via the {\tt ./quicksand} script in the top-level directory, which:
\begin{itemize}
	\item Checks if the user needs to run {\tt *-setup.sh} again, in case her source code was more recently updated than the existing annotated build (a common mistake),
	\item Passes its arguments through to {\tt id/landslide-id}, the Quicksand binary,
		and
	\item (If during the student user study,) compresses the resulting log files,
		creates a snapshot tarball of them and the current version of the user's code,
		and sends it to me for nefarious research purposes
		\revisionminor{(see \cref{sec:education-pebbles-instrumentation})}.
\end{itemize}

@@

\begin{itemize}
	\item {\tt -C}: enable ``control experiment'' mode, i.e., run only 1 instance of Landslide, with all (non-data-race) preemption points enabled in advance
	\item {\tt -I}: enable Iterative Context Bounding (requires {\tt -C}, although future work may relax this restriction);
		this generally causes bugs to be found faster should they exist, but degrades completion time
		(\cref{sec:landslide-icb})
	\item {\tt -0}: enable Preempt-Everywhere mode (\cref{sec:quicksand-eval}, requires {\tt -C})
	\item {\tt -M}: enable Maximal State Space mode,
		which prioritizes the maximal state space to optimize for fast verification,
		abandoning all subset jobs even if they might find bugs faster
		(\cref{sec:tm-eval}, incompatible with {\tt -C}).
		According to \cref{sec:quicksand-soundness}'s soundness proofs,
		this is equivalent to {\tt -0}
		(and according to my experience, \revisionminor{much} faster as well).
	\item {\tt -H}: use Limited Happens-Before for data-race analysis (\cref{sec:background-hb})
		(default for Pebbles kernelspace mode)
	\item {\tt -V}: use vector-clock-based Pure Happens-Before for data-race analysis (\cref{sec:background-hb})
		(default for P2/PSU userspace and Pintos modes)
	\item {\tt -X}: support transactional memory (\Cref{chap:tm})
	\item {\tt -A}: support multiple abort codes during transaction failure (\cref{sec:tm-implementation});
		required for testing programs which behave differently under different abort circumstances,
		but impacts the state space size
	\item {\tt -S}: suppress retry aborts during transaction failure (\cref{sec:tm-implementation})
	\item {\tt -R}: enable retry-set state space reduction for transactional tests (\cref{sec:tm-implementation})
	\item {\tt -P}: support Pintos architecture (enabled automatically when {\tt pintos-setup.sh} is run)
	\item {\tt -4}: support Pebbles architecture (enabled automatically when either {\tt p2-setup.sh} or {\tt psu-setup.sh} is run)
	\item {\tt -e ETAFACTOR}: configure heuristic state space ETA deferring factor (\cref{sec:quicksand-id}; described in detail in {\tt id/option.c})
	\item {\tt -E ETATHRESH}: configure heuristic threshold of state space progress for judging ETA stability (\cref{sec:quicksand-id}; described in detail in {\tt id/option.c})
\end{itemize}

@@

\begin{itemize}
	\item {\tt within\_function FUNC} - adds {\tt FUNC} to an allowlist of functions required to appear in the current stack trace before identifying a preemption point (see \cref{sec:landslide-pps})
	\item {\tt without\_function FUNC} - as above, but a denylist instead of an allowlist
	\item {\tt within\_user\_function FUNC} - as two above but finds the function in the userspace test program rather than the kernel code.
	\item {\tt without\_user\_function FUNC} - difference to two above same as stated one above.
	\item {\tt data\_race ADDR TID LAST\_CALL CURRENT\_SYSCALL} - specifies a data-race preemption point.
		\begin{itemize}
			\llitem {\tt ADDR} shall be the code address (in hex) of the racing address,
			{\em before} the execution of which a preemption will be issued.
			\llitem {\tt TID} indicates a thread ID required to be running for this data race.
				To specify data race PPs across all threads at once, set {\tt FILTER\_DRS\_BY\_TID=0} (see next section).
			\llitem {\tt LAST\_CALL} indicates a code address required to be the site of the last {\tt call} instruction executed
				(similar to specifying a stack trace, but using a full stack trace here degrades performance too much),
				or 0 to not use this feature.
				From personal experience I found this option rather useless and recommend always supplying 0.
				For further discussion see \cref{sec:quicksand-pps}.
			\llitem {\tt CURRENT\_SYSCALL} indicates the system call number if a user-space data race comes from within a kernel system call which accesses user memory (Pebbles only).
				Usually 0 (i.e., not in kernel code) but {\tt deschedule}'s system call number is common as well.
		\end{itemize}
	\item {\tt input\_pipe FILENAME} - FIFO file used for receiving messages from Quicksand (e.g. to suspend or resume execution).
		Requires {\tt id\_magic} option to be set (next section below).
		The odds that a human user will find spiritual enlightenment through using this option by hand are infinitesimal.
	\item {\tt output\_pipe FILENAME} - as above but for sending messages.
\end{itemize}

@@

\section{Kernel annotations}
\label{sec:landslide-kernel-anno}

@@

Hence,}
the educational experiments in this thesis focus on projects which students implement
on top of provided kernel basecode which Landslide already ``understands''.
Such understanding is conferred via the annotations described in this section.
For P2 and Pintos students I supply these annotations behind the scenes,
but a CMU 15-410 student who wishes to use Landslide on her kernel project shall need to brave forth hereupon.
An alternate step-by-step guide is also available at
\url{http://www.contrib.andrew.cmu.edu/~bblum/landslide-guide-kernel.pdf}.

@@

Traversal of the state space is implemented in three parts:
first, identifying preemption points when first encountered and selecting which thread to run for its first execution, in {\tt arbiter.c},
second, selecting which preemption point to backtrack to after completing an execution
and which thread to ``have switched to''
\footnote{\revision{willan on-having switched to \cite{hhgg-reu}}}
instead, in {\tt explore.c},
and third, rewinding the machine state to implement said backtracking,
in {\tt timetravel.c} (Bochs version) and {\tt timetravel-simics.c} (Simics version).

@@

The arbiter (named after the corresponding component of dBug \cite{dbug-ssv})
is responsible for checking which code locations during execution should be identified as preemption points
({\tt arbiter\_interested()}),
and thereupon for choosing whether to keep running the current thread or to preempt and switch to a new one
({\tt arbiter\_choose()}).
Its behaviour in the former case is configured by the options listed in \cref{sec:landslide-dynamicconfig},
and in the latter case by the options listed in \cref{sec:landslide-staticconfig}.
For example, {\tt EXPLORE\_BACKWARDS} is interpreted here;
if set, it will cause Landslide to always preempt and switch threads the first time it encounters each new preemption point.%
\footnote{Another secret option, {\tt CHOOSE\_RANDOMLY},
also exists here to randomize whether to ``explore backwards''
(choosing independently at each preemption point, resulting in an overall unpredictable exploration order).
It's not exposed to {\tt config.landslide} but rather the user must edit it in {\tt arbiter.c} directly,
whereupon the probability may also be adjusted via {\tt numerator} and {\tt denominator}.}

@@

The {\tt data\_race} command specifies an instruction pointer value to identify as a data race preemption point.
It can optionally be qualified by a thread ID, most recent system call number, et cetera,
as described in \cref{sec:landslide-dynamicconfig},
and is queried through {\tt suspected\_data\_race()}.

In Preempt-Everywhere mode, Landslide instead marks all shared memory accesses
as long as they are not part of either the mutex implementation or the running thread's stack frame.
The {\tt data\_race} command is ignored and {\tt suspected\_data\_race()}
instead checks whether the instruction pointer is associated with any such shared memory access.

@@

{\bf Choosing which new interleaving to test next.}
Among all interleavings chosen by DPOR not already marked as explored,
Landslide chooses the one with the longest execution prefix matching the current $S$,
to maintain the depth-first search invariant.
(In the case of a tie, differing only by which thread to run, it chooses arbitrarily.)
All other new interleavings are marked to explore later, %in their corresponding {\tt struct hax},
and automatically included in the result of any future iterations of DPOR until they are tested.
Because the one with the longest execution prefix was chosen to test next,
all others must share their execution prefixes with it,
preserving the $O(nk)$ memory bound described above.
%so storing markers for the others during the next execution is still $O(1)$ memory.

@@

In this interleaving, DPOR identifies 3 memory conflicts, two read/write and one write/write,
among the threads' 4 accesses to {\tt x}.
For each such pair, it ``marks'' an alternate interleaving,
which shall begin by preempting the thread of the first half of the conflict just before its execution thereof.
The ultimate goal is to execute an interleaving which reorders the conflict, which may expose new behaviour.
These marked interleavings form a work-queue which defines the exploration.
DPOR consumes from this set in depth-first order
(note the reversed order of $\bigstar$1 and \revisionminor{$\bigstar$3})
to avoid storing in memory any representation of exponentially-sized subtrees outside of the current branch.
Note also that in \revisionminor{$\bigstar$3},
the reordered \dporTB{{\tt tmp2~<-~x}} is not directly part of the memory conflict which marked it,
but it must be executed first to reach the conflicting \dporTB{{\tt x~<-~tmp2}}.

Now, let us run multiple iterations of DPOR
to advance through the first half of the full state space shown in \Cref{fig:tree}(b).
\Cref{fig:dpor-example-1} shows the outcome
%(with the previously-marked $\bigstar$2, now re-labeled as $\bigstar$3, having its subtree abbreviated). % old ver
\revision{(with the new $\bigstar$2 appearing in $\bigstar$1's subtree, to be explored before $\bigstar$3)}.

@@

@@

First of all, note that \Cref{alg:icb} is structured in a way that repeats interleavings
with fewer than $n$ preemptions that have already been checked in previous iterations of the outer loop.
This is because the number of preemptions in each branch is not known in advance;
rather, the state space must always be explored in an overall depth-first approach,
at best skipping too-preemptful interleavings as they are encountered.
As simple as it would be to state ``{\bf foreach} $S \in {\mathsf{sort}_P(U)}$'' in pseudocode,
implementing such an ordering would be much less straightforward.
\footnote{\revision{Many conference papers' pseudocode algorithms
often fail to be straightforward to translate into usable implementations~\cite[Figure 5]{itg2}.}}
